Comparison of several sequential and parallel derivative-free global optimization algorithms

### 2018-11-30 2019/PAVT/pavt_paper.tex
Some corrections. not final

\begin{abstract}
This work considers several stochastic and deterministic derivative-free global optimization
algorithms. In the first part of the paper popular sequential open-source solvers are compared against
Globalizer solver, which is developed at the Lobachevsky State University. The Globalizer is
designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. The second part of this work is devoted to
comparison between the Globalizer and MIDACO solvers on systems with shared and distributed
memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization
environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-
dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

Nonlinear global optimization of non-convex functions is traditionally considered to be one of the most
difficult problems of mathematical programming. The finding of the global minimum of
a function of several variables often appears to be more difficult than local optimization in the
thousand-dimensional space. In local optimization applying of the simplest gradient descent
method may appear to be enough, whereas in order to \textit{guarantee} find the global
optimum, an optimization method has to accumulate information about behavior of the
objective function within the whole search domain
\cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}. Nowadays, various stochastic global
optimization algorithms have attracted much attention, first of all, the evolution ones
\cite{Storn1997, SCHLUTER2009, KennedyEberhart1995}. They have rather simple structure
and allow solving the problems of large dimensionality. However, the evolution algorithms ensure
the global convergence in the probabilistic sense only.

@@

\subsection{Parallel Algorithm of Global Search}
\subsubsection{Dimension Reduction with Evolvents}
Within the framework of the information-statistical global optimization theory,
the Peano space-filling curves (or evolvents) \(y(x)\) mapping the interval \([0,1]\)
onto an \(N\)-dimensional hypercube \(D\) unambiguously are used for the dimensionality
reduction \cite{sergeyevStronginLera2013, strongin1978, strSergGO}.
\par
As a result of the reduction, the initial multidimensional global optimization
problem (\ref{eq:task}) is reduced to the following one-dimensional problem:
\begin{equation}
\label{eq:oneDimTask}
\varphi(y(x^*))=\min\{\varphi(y(x)):x\in [0,1]\}.
\end{equation}
\par
It is important to note that this dimensionality reduction scheme transforms the % minimized
Lipschitzian function from (\ref{eq:task}) to the corresponding one-dimensional
function \(\varphi(y(x))\), which satisfies the uniform H{\"o}lder condition, i. e.
\begin{equation}
\label{eq:holder}
|\varphi(y(x_1))-\varphi(y(x_2))|\leq H{|x_1-x_2|}^{\frac{1}{N}}, x_1,x_2\in[0,1],
\end{equation}
where the constant $H$ is defined by the relation \(H=2L\sqrt{N+3}\), \(L\) is the Lipschitz
constant from (\ref{eq:lip}), and \(N\) is the dimensionality of the optimization problem
(\ref{eq:task}).
\par
The algorithms for the numerical construction of the Peano curve approximations are
given in \cite{strSergGO}.

@@

\subsection{Parallel Algorithm on Distributed Memory Exploiting a Set of Evolvents}
\label{sub:parallel_evolvents}
\subsubsection{Rotated Evolvents}
One of the possible ways to overcome the negative effects of using a numerical
approximation of evolvent (it destroys the information about the neighbor points in
$\mathbb{R}^N$ space)
consists in using the multiple mappings
\begin{equation}
  Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{equation}
instead of single Peano curve $y(x)$ (see \cite{strSergGO}).
The building of a set of Peano curves by rotation of the evolvents around the coordinate origin is a
distinctive feature was proposed in \cite{Gergel2009}. Taking into account the initial mapping, one
can conclude that current implementation of the
method allows to build up to $N(N-1)+1$ evolvents for mapping the $N$-dimensional domain
onto the corresponding one-dimensional intervals. This method for building a set of mappings can be
''scaled'' easily to obtain more evolvents (up to
$2^N$) if necessary.

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leq s\leq S
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s^{th}\) problem, consists in the following sequence
of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leq s\leq S\), of the point \(y^k\) and
interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the
Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an
accessible core.
\end{itemize}

@@

  \item \textbf{Locally-based Algorithm of Global Search (AGS\(l\))} \cite{indexMethod}. It's a
modification of
  the original AGS which make it more locally oriented by alternately using of two types of
characteristics in the Rule 2 from the Section \ref{sub:ags}.

@@

  \item \textbf{Locally-based DIRECT (DIRECT$l$)} \cite{Gablonsky2001}. It's a variation of
DIRECT which pays less attention to non-promising boxes and therefore
  has less exploration power: it can converge faster on problems with few local minima, but lost the
global one in complicated cases.

@@

All the mentioned algorithms (except of AGS\(l\) which implemented in the Globalizer system)
are available in source codes as parts of wide-spread optimization packages.
AGS, DIRECT, DIRECT$l$, CRS, MLSL and StoGO are part of the NLOpt library \cite{nlopt}.
Differential Evolution and DSA can be found in
the latest version of the SciPy \cite{scipy} package for Python.

>>>>

\begin{abstract}
This work considers several stochastic and deterministic deri-vative-free global optimization
algorithms. In the first part of the paper popular sequential open-source solvers are compared against
Globalizer solver, which is developed at the Lobachevsky State University. The Globalizer is
designed to solve problems with black-box objectives satisfying the Lipschitz condition and shows
competitive performance with other similar solvers. The comparison is done on several sets of
challenging multi-extremal benchmark functions. The second part of this work is devoted to
comparison between the Globalizer and MIDACO solvers on systems with shared and distributed
memory. MIDACO is a state-of-the-art global solver included to the TOMLAB optimization
environment for MATLAB. Results of the benchmark show advantages of the Globalizer on small-dimensional, but sufficiently multi-extremal benchmark functions.
% We would like to encourage you to list your keywords within
% the abstract section using the \keywords{...} command.
\keywords{deterministic global optimization $\cdot$ stochastic global optimization
  $\cdot$ parallel numerical methods $\cdot$ derivative-free algorithms $\cdot$ black-box
optimization}
\end{abstract}
%
\section{Introduction}

Nonlinear global optimization of non-convex functions is considered to be one of the most
difficult problems of mathematical programming. The finding of the global minimum of
a function of several variables often appears to be more difficult than local optimization in a
thousand-dimensional space. In local optimization applying of the simplest gradient descent
method may appear to be enough, whereas in order to \textit{guarantee} finding the global
optimum, an optimization method has to accumulate information about behavior of the
objective function within the whole search domain
\cite{Jones2009,Paulavicius2011,Evtushenko2013,strSergGO}. Nowadays, various stochastic global
optimization algorithms have attracted much attention, first of all, the evolution ones
\cite{Storn1997, SCHLUTER2009, KennedyEberhart1995}. They have rather simple structure
and allow solving the problems of large dimensionality. However, the evolution algorithms ensure
the global convergence in the probabilistic sense only.

@@

\subsection{Parallel Algorithm of Global Search}
\subsubsection{Dimension Reduction with Space-Filling Curves}
Within the framework of the information-statistical global optimization theory,
the Peano space-filling curves (or \textit{evolvents}) \(y(x)\) mapping the interval \([0,1]\)
onto an \(N\)-dimensional hypercube \(D\) unambiguously are used for the dimensionality
reduction \cite{sergeyevStronginLera2013, strongin1978, strSergGO}.
\par
As a result of the reduction, the initial multidimensional global optimization
problem (\ref{eq:task}) is reduced to the following one-dimensional problem:
\begin{equation}
\label{eq:oneDimTask}
\varphi(y(x^*))=\min\{\varphi(y(x)):x\in [0,1]\}.
\end{equation}
\par
It is important to note that this dimensionality reduction scheme transforms the % minimized
Lipschitzian function from (\ref{eq:task}) to the corresponding one-dimensional
function \(\varphi(y(x))\), which satisfies the uniform H{\"o}lder condition, i. e.
\begin{equation}
\label{eq:holder}
|\varphi(y(x_1))-\varphi(y(x_2))|\leq H{|x_1-x_2|}^{\frac{1}{N}}, x_1,x_2\in[0,1],
\end{equation}
where the constant $H$ is defined by the relation \(H=2L\sqrt{N+3}\), \(L\) is the Lipschitz
constant from (\ref{eq:lip}), and \(N\) is the dimensionality of the optimization problem
(\ref{eq:task}).
\par
The algorithms for the numerical construction of the Peano curve approximations are
given in \cite{strSergGO}.

@@

\subsection{Parallel Algorithm on Distributed Memory Exploiting a Set of Evolvents}
\label{sub:parallel_evolvents}
\subsubsection{Rotated Evolvents}
One of the possible ways to overcome the negative effects of using a numerical
approximation of evolvent (it destroys the information about the neighbor points in
$\mathbb{R}^N$ space)
consists in using the multiple mappings
\begin{equation}
  Y_L(x)=\left\{y^0(x),\ y^1(x),...,\ y^L(x)\right\}
\end{equation}
instead of single Peano curve $y(x)$ (see \cite{strSergGO}).
The building of a set of Peano curves by rotation of the evolvents around the coordinate origin is a
distinctive feature was proposed in \cite{Gergel2009}. Taking into account the initial mapping, one
can conclude that current implementation of the
method allows to build up to $N(N-1)+1$ evolvents for mapping the $N$-dimensional domain
onto the corresponding one-dimensional intervals. This method for building a set of mappings can be
``scaled'' easily to obtain more evolvents (up to
$2^N$) if necessary.

Using the multiple mapping allows solving initial problem (\ref{eq:task}) by parallel solving the
problems
\[
\min\{\varphi(y^s(x)):x\in [0,1]\}, 1\leq s\leq S,
\]
on a set of intervals $[0,1]$ by the index method. Each one-dimensional problem is solved on a
separate processor. The trial results at the point \(x^k\) obtained for the problem being solved by
particular processor are interpreted as the results of the trials in the rest problems (in the
corresponding points \(x^{k_1},\dots,x^{k_S})\). In this approach, a trial at the point \(x^k \in
[0,1]\) executed in the framework of the \(s^{th}\) problem, consists in the following sequence
of
operations:
\par
1. Determine the image \(y^k=y^s (x^k)\) for the evolvent \(y^s (x)\).
\par
2. Inform the rest of processors about the start of the trial execution at the point \( y^k\) (the
blocking of the point \(y^k\) ).
\par
3. Determine the preimages \(x{}^{k_s}  \in [0,1], 1\leq s\leq S\), of the point \(y^k\) and
interpret the
trial executed at the point \(y^k \in D \) as the execution of the trials in the \(S\) points
\(x{}^{k_1} ,\dots,x{}^{k_s} \)
\par
4. Inform the rest of processors about the trial results at the point \(y^k\).
\par
The decision rules for the mentioned parallel algorithm, in general, are the same as the rules of the
sequential algorithm (except the method of the trial execution). Each processor has its own copy
of the software realizing the computations of the problem functions and the decision rule of the
index algorithm. For the organization of the interactions among the processors, the queues are
created on each processor, where the processors store the information on the executed iterations
in the form of the tuples: the processor number \(s\), the trial point \(x{}^{k_s}\).
The modifications of the method taking into account multiply criteria are given in \cite{GERGEL2017,Gergel2018}.
\par
The mentioned parallelization scheme was proposed in \cite{Gergel2009} and implemented in the
Globalizer system with the use of MPI technology. Main
features of implementation consist in the following:
\begin{itemize}
  \item A separate MPI-process is created for each
 of \(S\) one-dimensional problems being solved, usually, one process per one processor
 employed.
 \item Each process can use $p$ threads to parallel execute $p$ trials, usually one thread per an
accessible core.
\end{itemize}

@@

  \item \textbf{Locally-biased Algorithm of Global Search (AGS\(l\))} \cite{indexMethod}. It's a
modification of
  the original AGS which make it more locally oriented by alternately using of two types of
characteristics in the Rule 2 from the Section \ref{sub:ags}.

@@

  \item \textbf{Locally-biased DIRECT (DIRECT$l$)} \cite{Gablonsky2001}. It's a variation of
DIRECT which pays less attention to non-promising boxes and therefore
  has less exploration power: it can converge faster on problems with few local minima, but lost the
global one in complicated cases.

@@

All the mentioned algorithms are available in source codes as parts of wide-spread optimization packages (except of AGS and AGS\(l\), which are implemented in the Globalizer system).
DIRECT, DIRECT$l$, CRS, MLSL and StoGO are part of the NLOpt library \cite{nlopt}.
Differential Evolution and DSA can be found in
the latest version of the SciPy \cite{scipy} package for Python.
